---
title: "Calcualting the Cost of Severe Weather Events"
author: "Steve Rowe"
date: "June 20, 2015"
output: html_document
---

## Synopsis
Severe weather events can cost lives and cause economic damage.  The NOAA Storm Data database contains information about
the damage caused by various types of weather.  Based on analysis of this data, we can determine the type of events causing
the most damage and those harming the most people.  The analysis in this paper concludes that flooding causes the most economic damage and tornados kill and harm the most people. 




## Data Processing
Analysis starts with data from the National Weather Service [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).  We load the data into R and reduce the data
set to only those events which cause either economic or population damage.

```{r libraries, message=F}
library(dplyr)
library(ggplot2)
library(tidyr)
```

```{r loading, cache=TRUE}
StormData <- read.csv("C:/Users/strowe/Desktop/jhds/RepData_Project2/repdata-data-StormData.csv", stringsAsFactors = F)
StormData <- mutate(StormData, CROPDMGEXP = factor(CROPDMGEXP), PROPDMGEXP = factor(PROPDMGEXP))
```

Once the data is loaded, it must be cleaned up.  The data dates back to 1950 and several terms have been used over time for
the various events.  The following code removes events which are indeterminate in nature and consolidates others.

```{r cleanup}
# remove all non-damaging events
damage <- filter(StormData, FATALITIES != 0 | INJURIES != 0 | PROPDMG != 0 | CROPDMG != 0)

#remove items which are not events or are unidentifiable
damage <- damage %>% filter(EVTYPE !="?" & EVTYPE !="Other" & EVTYPE !="OTHER" & EVTYPE !="URBAN SMALL" & 
                            EVTYPE != "APACHE COUNTY" & EVTYPE != "HIGH" & EVTYPE != "TSTMW")

#consolidate events
damage <- mutate(damage, EVTYPE = ifelse(grepl("winter", EVTYPE, ignore.case=T), "Winter Weather", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("tornado|torndao|funnel|dustnado|dust devil|gustnado", EVTYPE, ignore.case=T), "Tornado", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("blizzard", EVTYPE, ignore.case=T), "Blizzard", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("winter storm|wintry", EVTYPE, ignore.case=T), "Winter Storm", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("thunder", EVTYPE, ignore.case=T), "Thunder Storm", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("wind", EVTYPE, ignore.case=T), "Wind", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("tropical", EVTYPE, ignore.case=T), "Tropical Storm", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("wind", EVTYPE, ignore.case=T), "High Wind", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("flood|high water|fld|rising water", EVTYPE, ignore.case=T), "Flooding", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("high tide|surge|swell|wave|surf", EVTYPE, ignore.case=T), "Storm Surge/Tide", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("snow", EVTYPE, ignore.case=T), "Snow", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("freezing rain", EVTYPE, ignore.case=T), "Freezing Rain", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("rain|precip|burst|shower", EVTYPE, ignore.case=T), "Rain", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("cold|hypothermia|low temp|freeze|exposure", EVTYPE, ignore.case=T), "Cold", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("fire", EVTYPE, ignore.case=T), "Wildfire", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("hail", EVTYPE, ignore.case=T), "Hail", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("heat|warm", EVTYPE, ignore.case=T), "Excessive Heat", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("ice|icy|frost|freezing drizzle|sleet", EVTYPE, ignore.case=T), "Ice", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("landslide|mud|landslump", EVTYPE, ignore.case=T), "Landslide", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("hurricane|typhoon|tsunami", EVTYPE, ignore.case=T), "Hurricane", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("fog", EVTYPE, ignore.case=T), "Fog", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("avalanche|rock slide|avalance", EVTYPE, ignore.case=T), "Avalanche", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("lightning|lighting|ligntning", EVTYPE, ignore.case=T), "Lightning", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("drown|Marine accident|marine mishap|rough seas|heavy seas", EVTYPE, ignore.case=T), "Marine Accident", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("coastal.*storm", EVTYPE, ignore.case=T), "Marine Accident", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("lightning", EVTYPE, ignore.case=T), "Lightning", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("waterspout", EVTYPE, ignore.case=T), "Waterspout", EVTYPE))
damage <- mutate(damage, EVTYPE = ifelse(grepl("rip", EVTYPE, ignore.case=T), "Rip Current", EVTYPE))
```
Next we must adjust the cost figures.  The database stores costs and an exponent which states the denomination of the 
cost.  For instance, a "B" would mean billions.
```{r econ}
# fix economic costs
damage$CROPDMG[damage$CROPDMGEXP=="B"]<- damage$CROPDMG[damage$CROPDMGEXP=="B"]*1000000000
damage$CROPDMG[damage$CROPDMGEXP=="m"]<- damage$CROPDMG[damage$CROPDMGEXP=="m"]*1000000
damage$CROPDMG[damage$CROPDMGEXP=="M"]<- damage$CROPDMG[damage$CROPDMGEXP=="M"]*1000000
damage$CROPDMG[damage$CROPDMGEXP=="k"]<- damage$CROPDMG[damage$CROPDMGEXP=="k"]*1000
damage$CROPDMG[damage$CROPDMGEXP=="K"]<- damage$CROPDMG[damage$CROPDMGEXP=="K"]*1000
damage$CROPDMG[damage$CROPDMGEXP=="h"]<- damage$CROPDMG[damage$CROPDMGEXP=="h"]*100
damage$CROPDMG[damage$CROPDMGEXP=="H"]<- damage$CROPDMG[damage$CROPDMGEXP=="H"]*100

damage$PROPDMG[damage$PROPDMGEXP=="B"]<- damage$PROPDMG[damage$PROPDMGEXP=="B"]*1000000000
damage$PROPDMG[damage$PROPDMGEXP=="m"]<- damage$PROPDMG[damage$PROPDMGEXP=="m"]*1000000
damage$PROPDMG[damage$PROPDMGEXP=="M"]<- damage$PROPDMG[damage$PROPDMGEXP=="M"]*1000000
damage$PROPDMG[damage$PROPDMGEXP=="k"]<- damage$PROPDMG[damage$PROPDMGEXP=="k"]*1000
damage$PROPDMG[damage$PROPDMGEXP=="K"]<- damage$PROPDMG[damage$PROPDMGEXP=="K"]*1000
damage$PROPDMG[damage$PROPDMGEXP=="h"]<- damage$PROPDMG[damage$PROPDMGEXP=="h"]*100
damage$PROPDMG[damage$PROPDMGEXP=="H"]<- damage$PROPDMG[damage$PROPDMGEXP=="H"]*100

```

## Results
There are two questions we are trying to answer: 

* Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
* Across the United States, which types of events have the greatest economic consequences?

### Population Health Effects
Population health effects includes both injuries and deaths.  The following chart indicates the total number of each for the 
various event types.  Tornados cause both the most injuries and the most death.  Heat, Flooding, and High Winds all stand
out as significant causes of injuries.

```{r popdmg }
popdmg <- damage %>% filter(FATALITIES != 0 | INJURIES != 0) %>% gather(type, count, FATALITIES:INJURIES) %>%
  group_by(EVTYPE, type) %>% summarize (total = sum(count))
ggplot(data=popdmg, aes(x=EVTYPE, y=total, fill=type)) + geom_bar(stat="identity") + coord_flip() +
  xlab("Event Type") +
  ylab("Total Injuries/Deaths") +
  ggtitle("Population Harm From Weather Events")
```

### Economic Consequences
Economic costs are expressed in terms of both crop damage and property damage.  The following chart indicates the total 
number of each for the various event types.  Flooding causes the most overall damage as well as the most property damage.
Drought causes the most crop damage.  Hurricanes, tornados, and storm surges also stand out as having high overall 
economic consequences.
``` {r econdmg, fig.height =6}
econdmg <- damage %>% filter(CROPDMG != 0 | PROPDMG != 0) 
econdmg <- gather(econdmg, type, count, c(CROPDMG, PROPDMG)) %>% group_by(EVTYPE, type) %>% summarize (total = sum(count))

ggplot(data=econdmg, aes(x=EVTYPE, y=total, fill=type)) + geom_bar(stat="identity") + coord_flip() +
  xlab("Event Type") +
  ylab("Total Cost") +
  ggtitle("Economic Damage From Weather Events")
```

